---
title: "Vatrix PCA (12G)"
format:
  html:
    embed-resources: true
editor: visual
---

## Introduction

This notebook loads the Vatrix MTX file on the 12G pool of nuclei to hopefully produce a PCA/UMAP with 8 clusters of nuclei.

## Libraries

Attach the packages that we need for the analysis.

```{r}
#| message: false
library(Matrix)
library(umap)
library(tidyverse)
```

```{r}
theme_set(theme_bw())
```

## Load data

```{r}
#| echo: false
matrix_alt <- Matrix::readMM("../results/vartrix/12G.alt.mtx")
str(matrix_alt)
```

```{r}
#| echo: false
matrix_ref <- Matrix::readMM("../results/vartrix/12G.ref.mtx")
str(matrix_ref)
```

## Basic filtering

Remove variants (i.e. rows) that are not detected in at least 10 nuclei. It's a small arbitrary filter but it'll get rid of the most useless rows.

```{r}
keep_variants <- rowSums(matrix_alt) >= 10 & rowSums(matrix_ref) >= 10
table(keep_variants)
```

```{r}
matrix_alt_filtered <- matrix_alt[keep_variants, ]
matrix_ref_filtered <- matrix_ref[keep_variants, ]
```

```{r}
alt_rank_stats <- 
  tibble(
    col_sum = colSums(matrix_alt_filtered)
  ) |> 
  mutate(
    rank = row_number(-col_sum)
  )
ref_rank_stats <- 
  tibble(
    col_sum = colSums(matrix_ref_filtered)
  ) |> 
  mutate(
    rank = row_number(-col_sum)
  )
```

```{r}
ggplot() +
  geom_point(aes(rank, col_sum), ref_rank_stats, colour = "blue") +
  geom_point(aes(rank, col_sum), alt_rank_stats, colour = "red")
```

```{r}
keep_nuclei <- colSums(matrix_alt_filtered > 0) & colSums(matrix_ref_filtered > 0)
table(keep_nuclei)
```

```{r}
matrix_alt_filtered <- matrix_alt_filtered[, keep_nuclei]
matrix_ref_filtered <- matrix_ref_filtered[, keep_nuclei]
```

```{r}
sprintf("%i variants", nrow(matrix_alt_filtered))
sprintf("%i nuclei", ncol(matrix_alt_filtered))
```

```{r}
matrix_combined <- rbind(matrix_alt_filtered, matrix_ref_filtered)
dim(matrix_combined)
```

## Dimensionality reduction

Run a quick and dirty PCA and UMAP to check whether the available data captures the expected number of samples.

```{r}
pca <- prcomp(x = t(matrix_combined), center = TRUE, scale. = FALSE)
str(pca)
```

```{r}
pca_stats <- tibble(
  PC = seq_along(pca$sdev),
  var_explained = pca$sdev^2 / sum(pca$sdev^2)
)
pca_stats
```

```{r}
ggplot(pca_stats, aes(PC, var_explained)) +
  geom_point()
```

```{r}
umap_out <- umap(d = pca$x[, 1:25])
umap_out
```

```{r}
umap_df <- as_tibble(as.data.frame(umap_out$layout))
colnames(umap_df) <- c("UMAP1", "UMAP2")
ggplot(umap_df, aes(UMAP1, UMAP2)) +
  geom_point()
```

