---
title: "report"
format:
  html:
    embed-resources: true
editor: visual
---

## Introduction

## Dependencies

```{r library}
#| message: false
library(BiocIO)
library(dplyr)
library(ggplot2)
library(readr)
library(rtracklayer)
library(DropletUtils)
```

## Helper functions

Load demuxlet data

```{r load_demuxlet}
load_demuxlet <- function(sample_id) {
  message("Loading demuxlet data for sample: ", sample_id)
  base_path <- "../results/popscle_demuxlet/%s.best"
  dexmulet_file <- sprintf(base_path, sample_id)
  demuxlet_data <- read_tsv(dexmulet_file, show_col_types = FALSE)
  demuxlet_data$sample_id <- sample_id
  demuxlet_data
}
# demuxlet_data <- load_demuxlet(sample_id = "A")
# demuxlet_data
```

Load cellranger filtered matrix

```{r load_cellranger_matrix}
load_cellranger_matrix <- function(sample_id) {
  message("Loading cellranger data for sample: ", sample_id)
  base_path <- "../../single_nuc_cellranger/results/cellranger/%s/outs/per_sample_outs/%s/count/sample_filtered_feature_bc_matrix/"
  cellranger_path <- sprintf(base_path, sample_id, sample_id)
  names(cellranger_path) <- sample_id
  sce <- read10xCounts(cellranger_path, col.names = TRUE)
  sce
}
# sce <- load_cellranger_matrix("A")
# sce
```

Add UMI sum to demuxlet data.

```{r}
# colData(sce)
# all(demuxlet_data$BARCODE %in% sce$Barcode)
# stopifnot(all(demuxlet_data$BARCODE == sce$Barcode))
# demuxlet_data$umi_sum <- colSums(counts(sce))
# demuxlet_data
```

## Load and combine results from all samples

```{r}
gtf_file <- "/ceph/project/tendonhca/shared/spatial/analysis/refdata-gex-GRCh38-2020-A/genes/genes.gtf"
gtf_gene_data <- import(gtf_file, feature.type = "gene")
mt_gene_ids <- mcols(subset(gtf_gene_data, seqnames == "chrM", "gene_id"))[["gene_id"]]
cat(mt_gene_ids, sep = "\n")
```

```{r load_sample_data}
load_sample_data <- function(sample_id) {
  demuxlet_data <- load_demuxlet(sample_id = sample_id)
  sce <- load_cellranger_matrix(sample_id)
  stopifnot(all(demuxlet_data$BARCODE == sce$Barcode))
  demuxlet_data$umi_sum <- colSums(counts(sce))
  demuxlet_data$n_features <- colSums(counts(sce) > 0)
  demuxlet_data$pct_mt <- colSums(counts(sce)[mt_gene_ids, ]) / demuxlet_data$umi_sum
  demuxlet_data
}
demuxlet_data <- load_sample_data("A")
demuxlet_data
```

Apply to all samples

```{r}
demuxlet_data <- do.call("rbind", lapply(c(
    "A",
    "B",
    "C",
    "D",
    "E",
    "F",
    "G",
    "H"), load_sample_data))
table(demuxlet_data$sample_id)
```

## Add ground truth

```{r ground_truth}
ground_truth <- tibble(
  sample_id = c(
    "A",
    "B",
    "C",
    "D",
    "E",
    "F",
    "G",
    "H"),
  sample_name = c(
    "LHBT_1502",
    "LHBT_1503",
    "LHBT_1506",
    "LHBT_1507",
    "LHBT_1519",
    "LHBT_1530",
    "LHBT_1532",
    "LHBT_1537"
    )
)
ground_truth
```

```{r add_ground_truth}
demuxlet_data <- demuxlet_data |> 
  left_join(ground_truth, by = "sample_id")
demuxlet_data
```

```{r prediction_type}
demuxlet_data$prediction_type <- factor(NA, c("AMB", "DBL", "SNG.correct", "SNG.wrong"))
demuxlet_data$prediction_type[demuxlet_data$DROPLET.TYPE == "AMB"] <- "AMB"
demuxlet_data$prediction_type[demuxlet_data$DROPLET.TYPE == "DBL"] <- "DBL"
demuxlet_data$prediction_type[demuxlet_data$DROPLET.TYPE == "SNG" & demuxlet_data$SNG.BEST.GUESS == demuxlet_data$sample_name] <- "SNG.correct"
demuxlet_data$prediction_type[demuxlet_data$DROPLET.TYPE == "SNG" & demuxlet_data$SNG.BEST.GUESS != demuxlet_data$sample_name] <- "SNG.wrong"
table(demuxlet_data$prediction_type, useNA = "ifany")
```

## Apply typical quality filters

```{r}
demuxlet_data <- demuxlet_data |> 
  filter(
    umi_sum >= 300 &
      n_features >= 300 &
      pct_mt <= 0.05
  )
```

## Analyse UMI sum relative to ground truth

```{r ggplot_prediction_type_umi_sum}
#| fig-width: 10
#| fig-height: 8
ggplot(demuxlet_data, aes(x = prediction_type, y = umi_sum)) +
  geom_jitter(width = 0.2, size = 0.1, colour = "grey") +
  geom_violin(fill = NA, scale = "width", draw_quantiles = 0.5) +
  facet_wrap(~sample_name) +
  scale_y_log10(limits = c(NA, NA)) +
  theme_classic()
```

Tabulate the numbers

```{r prediction_stats}
prediction_stats <- with(demuxlet_data, table(sample_name, prediction_type))
prediction_stats
```

Success rate?

```{r}
matrix(
  prediction_stats[,3] / rowSums(prediction_stats[,3:4]) * 100,
  ncol = 1,
  dimnames = list(NULL, "% correct SNG")
)
```

Minimum and median UMI sum for correctly assigned singlets.

```{r}
demuxlet_data |> 
  filter(DROPLET.TYPE == "SNG" & SNG.BEST.GUESS == sample_name) |> 
  group_by(sample_name) |> 
  summarise(
    n = n(),
    min_umi = min(umi_sum),
    median_umi = median(umi_sum)
  )
```
