---
title: "Report"
format:
  html:
    embed-resources: true
editor: visual
---

```{r sample_id}
sample_id <- "16G"
```

## Introduction

In this notebook, we'll load the CellRanger filtered matrix from the following file:

```{r inputfiles}
cellranger_filtered_matrix <- sprintf("/ceph/project/tendonhca/albrecht/pools_cellranger/results/cellranger/%s/outs/per_sample_outs/%s/count/sample_filtered_feature_bc_matrix", sample_id, sample_id)
names(cellranger_filtered_matrix) <- sample_id
```

## Libraries

What we need for this analysis:

```{r libraries}
#| echo: false
#| message: false
library(BiocIO)
library(dplyr)
library(DropletUtils)
library(forcats)
library(ggplot2)
library(readr)
library(rtracklayer)
library(Seurat)
library(scater)
library(scran)
library(scuttle)
library(tibble)
```

## Load data

Load the CellRanger filtered matrix.

```{r read10xcounts}
#| echo: false
sce <- read10xCounts(samples = cellranger_filtered_matrix, col.names = TRUE)
sce
```

## Apply typical quality filters

```{r}
gtf_file <- "/ceph/project/tendonhca/shared/spatial/analysis/refdata-gex-GRCh38-2020-A/genes/genes.gtf"
gtf_gene_data <- BiocIO::import(gtf_file, feature.type = "gene")
mt_gene_ids <- mcols(subset(gtf_gene_data, seqnames == "chrM", "gene_id"))[["gene_id"]]
cat(mt_gene_ids, sep = "\n")
```

```{r}
sce$umi_sum <- colSums(counts(sce))
sce$n_features <- colSums(counts(sce) > 0)
sce$pct_mt <- colSums(counts(sce)[mt_gene_ids, ]) / sce$umi_sum
keep <- sce$umi_sum >= 300 & sce$n_features >= 300 & sce$pct_mt <= 0.05
sce <- sce[, keep]
sce
```

## Basic processing

Adapted from <https://carpentries-incubator.github.io/bioc-scrnaseq/eda_qc.html>

Size factors:

```{r}
lib.sf <- librarySizeFactors(sce)
sf_df <- data.frame(size_factor = lib.sf)
ggplot(sf_df, aes(size_factor)) + 
  geom_histogram(fill = "grey", colour = "black") + 
  scale_x_log10() +
  labs(
    title = sprintf("Sample: %s", sample_id),
    subtitle = sprintf("Number of cells: %s", format(ncol(sce), big.mark = ","))
  ) +
  theme_bw()
```

Log-normalisation:

```{r}
sizeFactors(sce) <- lib.sf
sce <- logNormCounts(sce)
sce
```

Feature selection

```{r}
dec.sce <- modelGeneVar(sce)

fit.sce <- metadata(dec.sce)

mean_var_df <- data.frame(mean = fit.sce$mean,
                          var  = fit.sce$var)

ggplot(mean_var_df, aes(mean, var)) + 
    geom_point() + 
    geom_function(
      fun = fit.sce$trend, 
      color = "dodgerblue"
    ) + 
    labs(
      x = "Mean of log-expression",
      y = "Variance of log-expression"
    ) +
  theme_bw()
```

HVG

```{r}
hvg.sce.var <- getTopHVGs(dec.sce, n = 500)

mean_var_df$hvg <- rownames(mean_var_df) %in% hvg.sce.var
ggplot(mean_var_df, aes(mean, var, colour = hvg)) + 
    geom_point() + 
    geom_function(
      fun = fit.sce$trend, 
      color = "dodgerblue"
    ) + 
    labs(
      x = "Mean of log-expression",
      y = "Variance of log-expression"
    ) +
  theme_bw()

```

PCA

```{r}
sce <- runPCA(sce, subset_row = hvg.sce.var)
sce
```

PCA scree plot

```{r}
pct_var_df <- data.frame(
  PC = 1:50,
  pct_var = attr(reducedDim(sce), "percentVar")
)

ggplot(pct_var_df, aes(PC, pct_var)) + 
    geom_point() + 
    labs(y = "Variance explained (%)")
```

PCA scatter plot

```{r}
plotPCA(sce)
```

UMAP

```{r}
set.seed(111)

sce <- runUMAP(sce, dimred = "PCA")

plotUMAP(sce)
```

## Load deconvolution data

Load table

```{r}
demuxlet_file <- sprintf("/ceph/project/tendonhca/albrecht/pools_demuxlet/results/popscle_demuxlet/%s.best", sample_id)
demuxlet_data <- read_tsv(demuxlet_file, show_col_types = FALSE) |> 
  column_to_rownames("BARCODE")
stopifnot(all(colnames(sce) %in% rownames(demuxlet_data)))
demuxlet_data
```

Tabulate counts of droplets demultiplexed.

```{r}
table(demuxlet_data$DROPLET.TYPE, useNA = "always")
```

## Plot deconvolution results

Transfer deconvoluted labels to `sce` object.

```{r}
demuxlet_data$sce_label <- ifelse(demuxlet_data$DROPLET.TYPE == "SNG", demuxlet_data$SNG.BEST.GUESS, demuxlet_data$DROPLET.TYPE)
sce$demuxlet <- factor(
  demuxlet_data[colnames(sce), "sce_label"],
  c("AMB", "DBL", "LHBT_1519", "LHBT_1507", "LHBT_1537", "LHBT_1502",  "LHBT_1532", "LHBT_1503", "LHBT_1506", "LHBT_1530")
)
plotUMAP(sce, colour_by = "demuxlet")
```

UMI stats per droplet per donor:

```{r}
#| fig-width: 10
#| fig-height: 6
colData(sce) |> 
  ggplot(aes(x = demuxlet, y = umi_sum)) +
  geom_violin(scale = "width", draw_quantiles = c(0.5)) +
  geom_point(aes(x = demuxlet, y = umi_sum), colData(sce) |> as_tibble() |> filter(demuxlet == "LHBT_1532")) + # for pool 16G
  scale_y_log10(
    limits = c(300, 100E3)
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = sample_id,
    y = "Total UMI sum",
    x = "Assignment"
  ) +
  theme_classic()
```

Examine the number of SNPs used per droplet

```{r}
#| fig-width: 10
#| fig-height: 6
sce$demuxlet_snps <- demuxlet_data[colnames(sce), "NUM.SNPS"]
colData(sce) |> 
  ggplot(aes(x = demuxlet, y = demuxlet_snps)) +
  geom_violin(scale = "width", draw_quantiles = c(0.5)) +
  geom_point(aes(x = demuxlet, y = demuxlet_snps), colData(sce) |> as_tibble() |> filter(demuxlet == "LHBT_1532")) + # for pool 16G
  scale_y_log10(
    limits = c(1, 10E3)
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = sample_id,
    y = "Number of SNPs used by demuxlet",
    x = "Assignment"
  ) +
  theme_classic()
```

Examine the number of reads used per droplet

```{r}
#| fig-width: 10
#| fig-height: 6
sce$demuxlet_reads <- demuxlet_data[colnames(sce), "NUM.READS"]
colData(sce) |> 
  ggplot(aes(x = demuxlet, y = demuxlet_reads)) +
  geom_violin(scale = "width", draw_quantiles = c(0.5)) +
  geom_point(aes(x = demuxlet, y = demuxlet_reads), colData(sce) |> as_tibble() |> filter(demuxlet == "LHBT_1532")) + # for pool 16G
  scale_y_log10(
    limits = c(1, 10E3)
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = sample_id,
    y = "Number of UMIs used by demuxlet",
    x = "Assignment"
  ) +
  theme_classic()
```

Examine the fraction of UMIs used per droplet

```{r}
#| fig-width: 10
#| fig-height: 6
colData(sce) |> 
  ggplot(aes(x = demuxlet, y = demuxlet_reads / umi_sum)) +
  geom_violin(scale = "width", draw_quantiles = c(0.5)) +
  geom_point(aes(x = demuxlet, y = demuxlet_reads / umi_sum), colData(sce) |> as_tibble() |> filter(demuxlet == "LHBT_1532")) + # for pool 16G
  scale_y_continuous(
    limits = c(0, 1), labels = label_percent()
  ) +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = sample_id,
    y = "Fraction of UMIs used by demuxlet",
    x = "Assignment"
  ) +
  theme_classic()
```

Demonstrate that the number of SNPs and number of reads used by demuxlet are identical even if they are very similar and highly correlated:

```{r}
colData(sce) |> 
  as_tibble() |> 
  # filter(!demuxlet %in% c("AMB", "DBL")) |> 
  ggplot(aes(x = demuxlet_reads, y = demuxlet_snps)) +
  geom_abline(intercept = 0, slope = 1) +
  geom_point(aes(colour = demuxlet)) +
  scale_x_log10(limits = c(10, 100E3)) +
  scale_y_log10(limits = c(10, 100E3)) +
  scale_colour_discrete(drop = FALSE) +
  theme_classic() +
  labs(
    title = sample_id,
    x = "Number of reads used by demuxlet",
    y = "Number of SNPs used by demuxlet"
  )
```

Plot AMB, then singlets, then DBL:

```{r}
tmp_df <- cbind(
  colData(sce),
  reducedDim(sce, "UMAP")
) |> as_tibble()
ggplot() +
  geom_point(aes(x = UMAP1, y = UMAP2), tmp_df |> filter(demuxlet == "AMB"), colour = "grey") +
  geom_point(aes(x = UMAP1, y = UMAP2, colour = demuxlet), tmp_df |> filter(!demuxlet %in% c("AMB", "DBL"))) +
  geom_point(aes(x = UMAP1, y = UMAP2), tmp_df |> filter(demuxlet == "DBL"), colour = "red", shape = "x", size = 3) +
  guides(
    colour = guide_legend(override.aes = list(size = 3))
  ) +
  labs(
    title = sample_id
  ) +
  theme_classic()
```

Distribution of cell assignments:

```{r}
table(sce$demuxlet) |> as.data.frame()
```

Number of cells passed cellranger filtering AND deconvoluted:

```{r}
sum(!sce$demuxlet %in% c("AMB", "DBL"))
```

UMI count for cells, separated as singlet or not.

```{r}
#| fig-width: 4
#| fig-height: 5
sce$umi_sum <- colSums(counts(sce))
tmp_df <- colData(sce) |> 
  as_tibble() |> 
  mutate(
    singlet = !demuxlet %in% c("AMB", "DBL"),
    summary = fct_collapse(demuxlet, 
      AMB = c("AMB"), 
      DBL = c("DBL"), 
      other_level = "Singlet"
    )
  )
ggplot(tmp_df, aes(x = summary, y = umi_sum)) +
  geom_jitter(width = 0.2, colour = "grey") +
  geom_violin(fill = NA, scale = "width", draw_quantiles = 0.5) +
  scale_y_log10(limits = c(100, 50E3)) +
  labs(
    title = sample_id
  ) +
  theme_classic()
```

```{r}
n_assigned <- sum(demuxlet_data$DROPLET.TYPE == "SNG")
n_cellranger <- ncol(sce)
paste0(
  format(n_assigned, big.mark = ","),
  " assigned (",
  round(100* n_assigned / n_cellranger, digits = 1),
  "% of ",
  format(n_cellranger, big.mark = ","),
  " nuclei in cellranger matrix)"
)
```

```{r}
tmp_df <- tmp_df |> 
  arrange(desc(umi_sum)) |> 
  mutate(
    cumsum_singlet = cumsum(summary == "Singlet"),
    rank = row_number(),
    singlet_fraction = cumsum_singlet / rank
  )
tmp_df
```

```{r}
#| fig-width: 7
#| fig-height: 5
ggplot(tmp_df, aes(umi_sum, cumsum_singlet)) +
  geom_line() +
  scale_x_log10() +
  coord_cartesian(xlim = c(100, 1E5)) +
  labs(
    title = sample_id,
    y = "Total number assigned as singlet",
    x = "Minimum UMI sum"
  ) +
  theme_classic()
```

```{r}
#| fig-width: 7
#| fig-height: 5
ggplot(tmp_df, aes(umi_sum, singlet_fraction)) +
  geom_line() +
  scale_x_log10() +
  coord_cartesian(xlim = c(100, 1E5), ylim = c(0, 1)) +
  labs(
    title = sample_id,
    y = "Fraction assigned as singlet",
    x = "Minimum UMI sum"
  ) +
  theme_classic()
```

## Transfer learning

Carla's annotated object:

```{r}
carla <- readRDS("/project/tendonhca/shared/chromium/analysis/20231102_LHBT/20240116_15-16_Fine_annotation.dir/RDS_objects.dir/LHBT_integrated_annotated.rds")
carla
```

```{r}
DimPlot(carla, group.by = c("cell_annotation_update"))
```

```{r}
pool.query <- CreateSeuratObject(counts = counts(sce))
rownames(pool.query) <- make.unique(rowData(sce)[["Symbol"]])
keep_features <- intersect(rownames(carla), rownames(pool.query))
carla <- subset(carla, features = keep_features)
pool.query <- subset(pool.query, features = keep_features)
pool.query <- NormalizeData(pool.query)
pools.anchors <- FindTransferAnchors(
  reference = carla,
  query = pool.query,
  dims = 1:30,
  reference.reduction = "lognorm.pca"
)
predictions <- TransferData(
  anchorset = pools.anchors,
  refdata = carla$cell_annotation_update,
  dims = 1:30
)
pool.query <- AddMetaData(pool.query, metadata = predictions)
pool.query
```

```{r}
VlnPlot(pool.query, features = "prediction.score.max")  # or the metadata field name created by TransferData
```

```{r}
sce$predicted_cell_type <- pool.query$predicted.id
table(sce$predicted_cell_type)
```

```{r}
tmp_df <- cbind(
  colData(sce),
  reducedDim(sce, "UMAP")
) |> as_tibble()
n <- nlevels(carla$cell_annotation_update)
cols <- grDevices::hcl(
  h = seq(15, 375, length = n + 1)[1:n],
  l = 65,
  c = 100
)
names(cols) <- levels(carla$cell_annotation_update)
ggplot() +
  geom_point(
    aes(x = UMAP1, y = UMAP2),
    tmp_df |> filter(demuxlet == "AMB"),
    colour = "grey") +
  geom_point(
    aes(x = UMAP1, y = UMAP2, colour = predicted_cell_type),
    tmp_df |> filter(!demuxlet %in% c("AMB", "DBL"))
    ) +
  geom_point(
    aes(x = UMAP1, y = UMAP2), tmp_df |> filter(demuxlet == "DBL"),
    colour = "red",
    shape = "x",
    size = 3) +
  scale_colour_manual(values = cols) +
  guides(
    colour = guide_legend(override.aes = list(size = 3))
  ) +
  labs(
    title = sample_id
  ) +
  theme_classic()
```

```{r}
table(sce$predicted_cell_type, sce$demuxlet) |>
  as.data.frame() |> 
  pivot_wider(names_from = Var2, values_from = Freq) |> View()
```
